---
title: 'Case study: Viator inscriptions in viator'
author: 
- Petra Hermankova^[Aarhus University, Denmark, https://orcid.org/0000-0002-6349-0540]
date: "`r format(Sys.Date())`"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: true
    number_sections: true
    toc_depth: 3
    df_print: paged
---

_Purpose of this script is a basic exploration of the contents of viator inscriptions and its attributes, that contain reference to the term 'viator'._

# Initial setup

```{r setup, echo=TRUE, message=FALSE, warning = FALSE}

knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(tidyverse)
library(jsonlite)
library(tidytext)
library(reticulate)

```

```{r loading data}
# read in JSON
list_json <- jsonlite::fromJSON("2024-06-11-viator_via_Lat_Epig-term1_viator-697.json")
viator <- as_tibble(list_json$data)

# read in TSV
viator2 <- read_tsv("2024-06-11-viator_via_Lat_Epig-term1_viator-697.tsv")

```

# Preprocessing of the LatEpig dataset

## Type of inscription 

First we need to split the categorisation of the `status` attribute:

Division the status keywords into their own attributes, base on their purpose:
1. inscription type
2. notation of a societal status of a person on the inscription
3. details about process of inscribing, execution

```{r}
inscription_type_list <- c("tituli sepulcrales", "tituli fabricationis", "inscriptiones christianae", "tituli sacri", "tituli possessionis", "tituli operum", "miliaria", "tituli honorarii", "carmina", "signacula", "diplomata militaria", "leges", "defixiones", "termini", "reges", "signacula medicorum", "senatus consulta")

inscribing_process_list <- c("sigilla impressa", "litterae erasae", "litterae in litura", "tesserae nummulariae")

status_notation_list<- c("viri", "tria nomina", "mulieres", "nomen singulare", "liberti/libertae", "milites", "Augusti/Augustae", "ordo senatorius", "servi/servae", "officium/professio", "ordo decurionum", "sacerdotes pagani", "praenomen et nomen", "ordo equester", "seviri Augustales", "sacerdotes christiani")
```


```{r}
viator <- viator %>% 
  mutate(inscr_type = str_extract_all(pattern = paste(inscription_type_list, collapse="|"), string = viator$status)) %>% 
  mutate(status_notation = str_extract_all(pattern = paste(status_notation_list, collapse="|"), string = viator$status)) %>% 
  mutate(inscr_process = str_extract_all(pattern = paste(inscribing_process_list, collapse="|"), string = viator$status))
```

## Convert character (0) to NA

```{r}
viator$inscr_type <- lapply(viator$inscr_type, function(x) if(identical(x, character(0))) NA_character_ else x)
viator$status_notation <- lapply(viator$status_notation, function(x) if(identical(x, character(0))) NA_character_ else x)
viator$inscr_process <- lapply(viator$inscr_process, function(x) if(identical(x, character(0))) NA_character_ else x)
```

### Overview of types of inscriptions

```{r}
viator_insctype<- unnest_longer(viator, col = "inscr_type")
```

```{r}
viator_insctype %>% 
  count(inscr_type, sort=T)
```



# Text of an inscription 
- using the interpretive cleaned version
```{r}
# exploring the interpretive text of all inscriptions
text <- viator$inscription_interpretive_cleaning 

# displaying the first 6 records
head(text)
```

## Textual context

```{r}
# capturing context - 2 words before the term and 2 words after the term, including spaces, splitting lowercase viator from the personal name Viator
viator<- viator %>% 
  mutate(viator_in_context_10w = str_extract(viator$inscription_interpretive_cleaning, "(\\w*\\s){0,2}\\b[V|v]iator\\S{0,4}\\b(\\w*\\s){0,3}"))  %>% 
  mutate(viator_lowercase = str_extract(viator$inscription_interpretive_cleaning, "(\\w*\\s){0,2}\\bviator\\S{0,4}\\b(\\w*\\s){0,3}")) %>% 
  mutate(viator_uppercase = str_extract(viator$inscription_interpretive_cleaning, "(\\w*\\s){0,2}\\bViator\\S{0,4}\\b(\\w*\\s){0,3}"))
```

```{r}
# how many have viator in lowercase
length(na.omit(viator$viator_lowercase))

```

```{r}
# how many have Viator in uppercase, as a personal name 
length(na.omit(viator$viator_uppercase))

```

### Most common words
```{r}
context_tokenized<- viator %>% 
  unnest_tokens(word, inscription_interpretive_cleaning, token = stringr::str_split, pattern = " ") %>% 
  drop_na(word)

# this function converts all words to lowercase, thus the Viator and viator distinction is lost. That's why we distinguish them in the previous steps.
```

```{r}
context_tokenized %>% 
  count(word, sort = TRUE) %>% 
  head(20)
```

## Funerary inscriptions

```{r}
sepulcrales_lowercase<- viator_insctype %>% 
  filter(inscr_type == "tituli sepulcrales" & !is.na(viator_lowercase))

```

```{r}
# save as CSV for manual evaluation

write_csv(sepulcrales_lowercase, "LatEpig_viator_lowercase_tit_sepulcrales.csv")
```

### Manual evaluation / interpretation of the text

Upon close reading of the text of 226 inscriptions, we have noticed the following patterns:

1. `Viator` as passer-by, often addressed in adhortation to pause and learn about the date of the deceased, the most common verbs in imperative: `reste`(stop, pause), `siste` (stop), `consiste` (stop, stand still), `ave` (hail, farewell, salutation), `vale` (be well, saltutation), `scire` (know), `cognosce` (learn), in the first person singular `rogo te` (I ask you), or in plural imperative `avete vos viatores` (be well off travellers).

2. `Viator` as a low-ranking official, part of the `cursus honorum` of the person mentioned in the texts as an attendant of Roman magistrates, either as a noun or adjective: `viator tribunicius`, `viator consulum / consularis`, `viator quaestorius`, `viatori apparitori`, `viatores collegii`, `decurialis viatoris`.

3. Uncertain meaning due to the fragmentary preservation of the text

```{r}
# reading in manually evaluated dataset
sepulcrales_eval<- read_csv("LatEpig_viator_lowercase_tit_sepulcrales_eval.csv")

```


```{r}
# the ratio of the three groups

sepulcrales_eval %>% 
  count(viator_interpretation_manual, sort = T) %>% 
  mutate(ratio = n/(sum(n)/100))
```

### Bigrams of sepulcral inscriptions

To confirm the results of the close reading, we employ methods of distant reading, such as analysis of bigrams:

```{r}
sepulcrales_bigram <- sepulcrales_eval %>% 
  unnest_tokens(bigram, inscription_interpretive_cleaning , token = "ngrams", n = 2)
```

```{r}
# selecting bigrams that contain the pattern viator

sepulcrales_bigram %>% 
  count(bigram, sort = TRUE) %>% 
  filter(bigram %in% str_subset(bigram, pattern = "viator"))
```

```{r}
# overview of the manual evaluation and related bigrams

sepulcrales_bigram %>% 
  filter(bigram %in% str_subset(bigram, pattern = "viator")) %>% 
  count(viator_interpretation_manual, bigram, sort = TRUE)
```

### Time dimension
```{r}
# selecting only those with a date

sepulcrales_date <- sepulcrales_eval %>% 
  select(date_not_before, date_not_after, viator_interpretation_manual, latitude, longitude) %>% 
  filter(!is.na(date_not_before))
```

```{python}

# using the tempun from Python package to plot the date intervals of the funerary inscriptions

import geopandas as gpd # https://vincent.doba.fr/posts/20210407_install-fiona-on-windows/ & https://geopandas.org/getting_started/install.html Issues with geopandas on windows. The requirements.txt file might be able to install the packages, if not you should look into these geopandas links.
import requests
import seaborn as sns
import matplotlib.colors as mcolors
import matplotlib.pyplot as plt
from scipy.stats import trapz
import scipy
#import nltk
import json
import tempun
#import sddk
import numpy as np
import pandas as pd
pd.options.display.max_columns = 1000  # to see all columns
```


```{python}
# How many inscriptions have both dates?
sepul_date = r.sepulcrales_date[(r.sepulcrales_date["date_not_before"].notnull()) | (r.sepulcrales_date["date_not_after"].notnull())]
len(sepul_date)
```

```{python}
# Generate a list of 1000 random dates for each inscription in the dataset
sepul_date["random_dates"] = sepul_date.apply(lambda row: tempun.model_date(
    row["date_not_before"], row["date_not_after"], size=1000,), axis=1)
```

```{python}
# Count random dates in temporal timeblocks in a specified time range (from 200 BC to 600 AD, temporal blocks by 25 years)
sepul_complete_simulations = tempun.timeblocks_from_randoms(
    sepul_date["random_dates"], [-200, 600, 25])
```

```{python}
len(sepul_complete_simulations)
```

The following figures makes more sense if you are adding more than one dataset to them. This document is just a sample on how it can be done in RStudio, so it only uses one dataset, but it still creates an overview of how the output graphs could look.

```{python}
# Figure 1
# plot all the dataset to see the epigraphic production in time

plt.style.use("seaborn-white")
plt.rcParams['figure.dpi'] = 300
plt.rcParams['font.size'] = 10

tab_colors_list = list(mcolors.TABLEAU_COLORS.keys())

fig, ax = plt.subplots(figsize=(14, 6), tight_layout=True)

tempun.plot_timeblocks_data(
    sepul_complete_simulations, ax=ax, color="blue", label=f"Viator inscriptions (n={len(sepul_date)})")
    
ax.set_xlabel("Year", weight="bold")
ax.set_ylabel("Count", weight="bold")
ax.set_title("Viator inscriptions over time (MonteCarlo simulation, N=1000)")
ax.legend(title="EDCS / LatEpig inscriptions", title_fontsize="large", bbox_to_anchor=(1, 1), loc='upper right')

#plt.axvline(x=212, linewidth=0.5, color = 'orange', linestyle='dashed')

#fig.suptitle(f'Comparison of epigraphic production over time', fontsize=16,fontweight="bold")
plt.savefig('Fig1_viator_over_time.png')

fig
```


```{python}
# Inscriptions by meaning of the viator term

simulations_by_type_len_sepul = []

for ins_type in r.sepulcrales_date["viator_interpretation_manual"].unique():
    if ins_type != "NULL":
        subset_df = sepul_date[sepul_date["viator_interpretation_manual"] == ins_type]
        simulations_by_type_len_sepul.append((ins_type, len(subset_df)))
simulations_by_type_len_sepul = sorted(
    simulations_by_type_len_sepul, key=lambda x: x[1], reverse=True)
simulations_by_type_len_sepul

simulations_by_type_sepul = []

for ins_type_tup in simulations_by_type_len_sepul[:3]:
    subset_df = sepul_date[sepul_date["viator_interpretation_manual"]
                          == ins_type_tup[0]]
    simulations = tempun.timeblocks_from_randoms(
        subset_df["random_dates"], [-200, 600, 25])
    ins_type_N = len(subset_df)
    simulations_by_type_sepul.append([ins_type_tup[0], ins_type_N, simulations])
simulations_by_type_sepul = sorted(
    simulations_by_type_sepul, key=lambda x: x[1], reverse=True)

date_vars_by_instypes = []
for ins_type_tup in simulations_by_type_len_sepul[:10]:
    subset_df = sepul_date[sepul_date["viator_interpretation_manual"]
                          == ins_type_tup[0]]
    date_vars = []
    for n in range(100):
        date_vars.append(
            [date_var[n] for date_var in subset_df["random_dates"] if date_var != None])
    date_vars_by_instypes.append(date_vars)
```

```{python}
simulations_by_type_len_sepul
```

```{python}
# Figure 2

plt.style.use("seaborn-white")
plt.rcParams['figure.dpi'] = 300
plt.rcParams['font.size'] = 10

fig, ax = plt.subplots(figsize=(14, 6), tight_layout=True)
#fig, [ax1, ax2] = plt.subplots(2, 1, figsize=(14, 12), tight_layout=True)

for ins_type_sims, color in zip(simulations_by_type_sepul, tab_colors_list[:len(simulations_by_type_sepul)]):
    tempun.plot_timeblocks_data(ins_type_sims[2], ax=ax, color=color)

ax.set_xlabel("Year", weight="bold")
ax.set_ylabel("Count", weight="bold")
ax.set_title(f"Viator inscriptions by context over time (EDCS/LatEpig, n={len(sepul_date)})", weight="bold")

markers = [plt.Line2D([0, 0], [0, 0], color=color, lw=4)
           for color in tab_colors_list[:len(simulations_by_type_sepul)]]

legend_labels_sepul = [tup[0] + " (n={})".format(str(tup[1])) for tup in simulations_by_type_sepul]

ax.legend(markers, legend_labels_sepul, numpoints=1,  
           title=f"Viator group by context (n={len(sepul_date)})", title_fontsize="large", bbox_to_anchor=(1, 1), loc='upper right')

plt.savefig('Fig2_viator_typologies_comparison_over_time.png')
fig
```

### Plotting on a map

Here we are adding some definitions of the periods
```{python}
# periods definitions

periods = {  # to avoid overlaps, startdates are postponed by one year, when needed
    "Late Roman Republic": {"startdate": -200, "enddate": -26, "duration": 175},
    "Julio-Claudian dyn.": {"startdate": -27, "enddate": 68, "duration": 95},
    "Flavian dyn.": {"startdate": 69, "enddate": 96,"duration": 28},
    "Nerva-Antonine dyn.": {"startdate": 97, "enddate": 192, "duration": 96},
    "Severan dyn.": {"startdate": 193, "enddate": 235, "duration": 43},
    "Military emperors": {"startdate": 236, "enddate": 284,"duration": 49},
    "Tetrarchy-Constantine I": {"startdate": 285, "enddate": 337,"duration": 53},
    "Late Roman Empire": {"startdate": 338, "enddate": 600,"duration": 263}
}
timeblocks_periods = [(periods[period]["startdate"],
                       periods[period]["enddate"],
                       periods[period]["duration"]) for period in periods]
timeblocks_periods
```

```{python}
def date_to_str(date):
    if date < 0:
        date = str(abs(date)) + " BC"
    else:
        date = "AD " + str(date)
    return date


periods_labels = []
for period in periods.keys():
    start = date_to_str(periods[period]["startdate"])
    end = date_to_str(periods[period]["enddate"])
    periods_labels.append(period + "\n({0}-{1})".format(start, end))
periods_labels
```

In the following chunk we are loading a shapefile, that represents the Roman Empire at its largest extend in AD 117. This is used in the third figure. 
```{python}
# loading shapefile from Pleaides for the largest extent of the Roman Empire, AD 117
# source: https://raw.githubusercontent.com/pelagios/magis-pleiades-regions/main/pleiades-regions-magis-pelagios.geojson
pleiades_regions = gpd.read_file('pleiades-regions-magis-pelagios.geojson', driver='GeoJSON')
RE_merged = pleiades_regions.unary_union
```

```{python}
def get_date_var(randoms):
    try:
        return randoms[0]
    except:
        return None

# selecting one random date out of the 1000 version
sepul_date["date_var_1"] = sepul_date["random_dates"].apply(get_date_var)
```


```{python}
sepul_date_pb = sepul_date[sepul_date["viator_interpretation_manual"] == "passer-by"]
sepul_date_of = sepul_date[sepul_date["viator_interpretation_manual"] == "official"]
len(sepul_date_pb)
len(sepul_date_of)
```

```{python}
sepul_date_df_by_
```

```{python}
# passer-by
sepul_date_df_by_periods_pb = []

for period in timeblocks_periods:
    sepul_date_sample_pb = sepul_date_pb[sepul_date_pb["date_var_1"].between(
        period[0], period[1])]
    # tranforming data as geodataframe
    sepul_date_sample_pb = sepul_date_sample_pb[sepul_date_sample_pb["latitude"].notnull()]
    sepul_date_sample_pb = gpd.GeoDataFrame(sepul_date_sample_pb, geometry=gpd.points_from_xy(sepul_date_sample_pb["longitude"], sepul_date_sample_pb["latitude"]))
    # selecting only those dated and within borders of the Empire (the largest extent in AD 117)
    sepul_date_sample_pb = sepul_date_sample_pb[sepul_date_sample_pb.within(RE_merged)]
    sepul_date_df_by_periods_pb.append(sepul_date_sample_pb)
```

```{python}
len(sepul_date_df_by_periods_pb)
```

```{python}
# official
sepul_date_df_by_periods_of = []

for period in timeblocks_periods:
    sepul_date_sample_of = sepul_date_of[sepul_date_of["date_var_1"].between(
        period[0], period[1])]
    # tranforming data as geodataframe
    sepul_date_sample_of = sepul_date_sample_of[sepul_date_sample_of["latitude"].notnull()]
    sepul_date_sample_of = gpd.GeoDataFrame(sepul_date_sample_of, geometry=gpd.points_from_xy(sepul_date_sample_of["longitude"], sepul_date_sample_of["latitude"]))
    # selecting only those dated and within borders of the Empire (the largest extent in AD 117)
    sepul_date_sample_of = sepul_date_sample_of[sepul_date_sample_of.within(RE_merged)]
    sepul_date_df_by_periods_of.append(sepul_date_sample_of)
```


```{python}
# Figure 3
# plot all viator inscriptions with location in 8 plots grouped by period
plt.style.use("seaborn-white")
plt.rcParams['figure.dpi'] = 300
plt.rcParams['font.size'] = 10


fig, axs  = plt.subplots(4, 2, figsize=(7, 10), tight_layout=True)

contexts_pct = {}
    
for df_sepul_pb, df_sepul_of, ax, period in zip(sepul_date_df_by_periods_pb, sepul_date_df_by_periods_of, axs.ravel(), periods_labels):
    pleiades_regions.plot(ax=ax, color="lightgray")
    df_sepul_pb.plot(markersize=2, color="darkblue", ax=ax, alpha=0.9, label=len(df_sepul_pb))
    df_sepul_of.plot(markersize=2, color="red", ax=ax, alpha=0.9, label=len(df_sepul_of))
    ax.set_title(period, fontsize=6)
    ax.set_axis_off()
    markers = [plt.Line2D([0,0],[0,0],color=color, marker=".", linestyle="") for color in ["darkblue", "red"]]
    legend_labels = ["passer-by (n={0})".format(str(len(df_sepul_pb))),
                    "official (n={0})".format(str(len(df_sepul_of)))]
    
    ax.legend(markers, legend_labels, numpoints=1, bbox_to_anchor=(0.6, 1), loc='upper left', fontsize=6)
   
plt.tight_layout(pad=0.10)
plt.subplots_adjust(wspace=0, hspace=0.0)

fig.title(f'Spatial distribution of viator inscriptions by historic period', loc='center', fontsize=12, fontweight="bold")
plt.savefig('Fig3_viator_spatial_by_periods.png')
fig
```

```{python}
sepul_date_df_by_periods["viator_interpretation_manual"]
```



